Numerical modelling of the Cordilleran ice sheet

### 2014-07-20 kappa.tex
Add corrections from Arjen.

The North Americam Cordillera is an immense maze of mountain ranges and
intermontane basins which once hosted an ice mass of the size of Greenland: the
Cordilleran ice sheet. Although less extensive than the former Laurentide and
Fennoscandian ice sheets, the Cordilleran ice sheet perhaps remains the least
understood Pleistocene ice sheet in terms of its extent, volume, and dynamics.
Because of the irregular topography of the ice sheet's subglacial landscape,
ice flowed along tortuous pathways, sometimes following the major topographical
troughs, and sometimes flowing across them.

@@

The simulations presented in this thesis demonstrate the potential of combining
numerical ice sheet modelling to analysis of geological evidence in order to
improve our understanding former glaciations.

@@

Nevertheless, the landform record is sparse in time and, most often, spatially
incomplete. Palaeo-ice sheets did not leave a continuous imprint on the
lanscape, and much of this evidence has been overridden by subsequent glacier
re-advances and other geomorphological processes \citep{Kleman.1994,
Kleman.etal.2006, Kleman.etal.2010}.
Therefore, geomorphological reconstructions of former glacier extent require a
substantial amount of interpretation. This is where numerical ice sheet models,
incorporating the processes of glacier sliding \citep[e.g.,][]{Weertman.1957}
and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can
help. Although these models alone contain too many uncertainties to reproduce
accurate ice sheet reconstructions during former glacial cycles
\citep[e.g.,][]{Hebeler.etal.2008}, they can be calibrated against geological
data where they exist \citep{Napieralski.etal.2007}, while allowing
physics-based, quantitative reconstructions of ice sheet sectors where they do
not
\citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.
In addition, numerical ice sheet models provide quantitative insights on the
four-dimensional distribution of stresses and deformations in palaeo-ice sheets
and allow for reconstructions their past thickness and volume evolution, which
are difficult to infer from geological evidence\renote{
    Is this the place where to cite \citep{Denton.Hughes.1981}? \todo{I need to
    check the contents of this book (library).}
}.

@@

The \emph{dominant reliefs} occur along two major ribbons that run parallel to
each other along the general orientation of the North American Cordillera. The
western ribbon is formed by the Pacific Coast orogenic belt and includes, from
North to South, the Alaska Range (\cref{fig:photo-alaska-range}), the Wrangell
and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern ribbon consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench depression, and includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky mountains in the South. The
Brooks Range of northern Alaska has been covered by ice, but its glaciers
remained disjunct from the main body of the Cordilleran ice sheet
\citep{Kaufman.Manley.2004}.

@@

Evidence from glaciofluvial deposits dated using cosmogenic nuclides at
$2.64^{+0.20}/_{-0.18}$\,Ma indicates the first presence of an ice sheet in the
Cordilleran mountains during the early Pleistocene \citep{Hidy.etal.2013}.

This indicates that the North American Cordillera has been glaciated many times
since the late Pliocene, with the first of these glaciations occurring even
before the mid-Pleistocene onset of 100 ka glacial cycles
(\cref{fig:paleo-timeseries}). However, only fragments of the last few
glaciations can be reconstructed from the geologic record, implying that much
of the older evidence has been erased by subsequent glacial re-advances and
perhaps even by other, non-glacial geomorphological processes.

@@

Apart from geological evidence of the final demise of the ice sheet, most field
evidence relate to the LGM extent. It indicates that ice covered most of the
Canadian Cordillera, and some of the adjacent regions in north-western
contiguous USA and Alaska (\cref{fig:map-cordillera}). The Cordilleran ice
sheet originated from the coalescence of several mountain-centred ice caps
\citep{Davis.Mathews.1944}. It is generally thought that this LGM configuration
was short-lived \citep{Clague.etal.1980, Clague.1985, Cosma.etal.2008}, a
hypothesis which can be studied through numerical modelling.

The \emph{southern margin} of the ice sheet has been studied extensively, and
is in several respects better understood than its northern counterpart
\citep{Booth.etal.2003}. It consists of coalescent piedmont lobes, each of
which formed at the outlet of the main valleys. The Puget Lobe
\citep{Thorson.1980, Porter.Swanson.1998} was the westernmost and largest one.
It covered the Puget Lowland, a region that comprises the cities of Vancouver
and Seattle. Further east, beyond the Cascade Range, the Okanagan and Purcell
lobes formed the southern Cordilleran ice sheet margin. The latter once dammed
the Clark Fork River, inhibiting runoff and forming Glacial Lake Missoula
\citep{Pardee.1910}. As the ice sheet margin retreated, the lake repeatedly and
catastrophically drained to the Pacific Ocean \citep{Bretz.1923,Waitt.1980}.
These floods contributed to shape a distinctive landscape, known as the
Channeled Scabland.

The \emph{western margin} of the Cordilleran ice sheet covered part of the
Pacific continental shelf. An independent ice cap covering Vancouver Island
diverted flow from the main body of ice to drain into the Strait of Georgia
between Vancouver Island and the Olympic Peninsula, where it formed the Juan
the Fuca marine ice lobe, calving into the Pacific Ocean \mref. Further to the
North, the ice margin encroached the Queen Charlotte Islands. However, the
archipelago may not have been fully ice-covered, even during the maximum of
glaciation. Submarine troughs traversing several portions of the continental
shelf indicate the erosional action of additional ice streams, yet it is not
clear whether they have formed during the last glacial cycle or during earlier
glaciations \mref.

@@

The retreat of the ice sheet during the last \emph{deglaciation} was likely
rapid. Most of the evidence for the style and pace of deglaciation comes from
the Fraser Plateau. Studies there have indicated that some of the highest
reliefs emerged from the ice sheet prior to the surrounding lowlands, possibly
resulting in thin, stagnant ice remnants covering parts of the plateau
\citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}. Less
is known about the deglaciation of the northern sector of the Cordilleran ice
sheet, but abundant glacial landforms from the Liard Lowland indicate a
different mode of deglaciation, with active retreat of the ice margin towards
the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in
\CCYC\renote{
    I have not had time to dig into new literature. The Ward paper on
    deglacial valley fills in the Pelly River is like Chinese to me, and I
    could not access the Dyke paper on the Liard Lowland, assuming this is the
    one cited by Martin in his paper.}.

@@

@@

The thermodynamic core of the ice sheet model relies on four elementary field
equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the ice
density, $\rho$, is constant, and
conservation of mass therefore implies a conservation of volume, which can be expressed
in terms of the ice velocity vector, $\vv$ (\cref{fig:model-variables}), by
\begin{equation}
    \label{eqn:incompressibility}
    \div{\vv} = 0 \,.
\end{equation}

@@

Thirdly, after defining the strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$\renote{
  $\grad$, $\div$, $\tgrad$, $\tdiv$, are the classic gradient and divergence operators, applied on vector and tensor fields. $\tr$ is the trace of a matrix. I don't think these need to be defined.},
and the deviatoric stress tensor, ${\DST = \CST + p\,\IDT}$,
where $p=-\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
\citep{Nye.1953} based on laboratory experiments \citep{Glen.1952} is used to
relate these two quantities. It reads:
\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST \,.
\end{equation}
The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$, and
water content, $\omega$, through an Arrhenius-type law
\citep[Eqs.~63--65]{Paterson.Budd.1982, Aschwanden.etal.2012},
\begin{equation}
    A =
    \begin{cases}
        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
            & \text{if}\ T < T_c \,, \\
        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
            & \text{if}\ T \ge T_c \,,
    \end{cases}
\end{equation}
where $T_{pa}$ is the pressure-adjusted temperature calculated using the
Clapeyron relation, ${T_{pa} = T - \beta p}$. $R$ is the ideal gas constant,
and $A_c$, $A_w$, $Q_c$ and $Q_w$, are constant parameter corresponding to
values measured above and below a critical temperature threshold
${T_c=-10}$\,\degC\ \citep[\cref{tab:iceflowparams};][]{Paterson.Budd.1982}.
The water fraction, $\omega$, is capped at a maximum value of 0.01, above which
no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
Greve.1997}.

For a convenient derivation of the shallow shelf approximation, developed in
the next section (\cref{sec:siassa}), the constitutive law can be rewritten
using an inverted formulation. Assuming
${{\doteps_e}^2 = \frac{1}{2} \tr(\SRT^2)}$ and ${B=A^{-1/n}}$,
\cref{eqn:glenslaw} becomes
\begin{equation}
    \label{eqn:invglenslaw}
    \DST = B\,\doteps_e^{1/n-1}\,\SRT \,.
\end{equation}

By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
by ${\DST = 2 \nu\SRT}$, thus yielding
\begin{equation}
    \label{eqn:viscosity}
    \nu = \frac{B}{2}\,\doteps_e^{1/n-1} \,.
\end{equation}

@@

From \cref{eqn:vsia} it can be seen, that the shallow ice velocities, $\vsia$,
can be directly computed from the geometry of the ice sheet surface, $s$,
after a single vertical integration over the ice thickness, in order to
account for variations of ice rheology, $A$, with depth.

@@

In PISM, the SIA and SSA velocities are eventually added
\citep[\cref{fig:model-siassa};][Eq.~15]{Winkelmann.etal.2011}:
\begin{equation}
    \label{eqn:siassa}
    \vv = \vsia + \vssa \,.
\end{equation}
Because $\vsia$ is small where $\vssa$ is large and
reciprocally, we can expect the model to perform well in regions where
conditions for applicability of either the SIA or the SSA are met
(\cref{fig:model-siassa}). However,
\cref{eqn:siassa} is a heuristic, and its validity in the zone of
transition between shear flow (SIA) and longitudinal flow (SSA) has not been
mathematically proven.

@@

The till cohesion $c_0$ is assumed to be zero\renote{
  Yes, we should probably defend this choice, but I am not aware of any
  observations justifying this.}.
The effective pressure on the
till, $N$, is determined by the modelled amount of water at the bed. However,
two different parametrisations are used within this thesis. In {\CCLI}, using
PISM~0.5, the effective pressure is linearly related to water content by a
simple parametrisation \citep[Eq.~13]{Winkelmann.etal.2011},
\begin{equation}
    N = \rho gh (1 - \alpha \frac{W}{W_{max}}) \,,
\end{equation}
where $W$ is the amount of water at the bed, and varies from zero to
$W_{max}=2$\,m, a threshold above which further melt simply exits the system,
thus assuming instantaneous drainage. Although mass-conserving water drainage
is now implemented in PISM \citep{Bueler.Pelt.2014}, is not included in the
simulations presented here for the sake of simplicity.

@@

The modelled ice sheet interacts with the underlying bedrock in two ways.
Firstly, \emph{bedrock temperatures} are computed to a given depth by
appending the
ice enthalpy model (\ref{eqn:enthalpy}) with an underlying bedrock thermal
model. The only process accounted for is vertical heat conduction, hence this model
consists of a simple diffusion equation,
\begin{equation}
    \pdv{T}{t} = \frac{k_b}{\rho_b c_b} \Delta T \,,
\end{equation}
where $k_b$, $\rho_b$ and $c_b$ are the thermal conductivity, density and
specific heat capacity of the bedrock (\cref{tab:bedrockparams}).
Bedrock temperature is conditioned at depth by a fixed upwards geothermal heat
flux $q_G$  (\cref{tab:bedrockparams}). The bedrock thermal model
is necessary for modelling the Cordilleran ice sheet on multi-millennial
time-scales, because temperature fluctuations caused by climate change and
varying isolation from the fluctuating ice sheet penetrate several kilometres
into the rock, and eventually affect basal ice temperatures.

Secondly, bedrock elevation evolve in response to the ice weight. The
\emph{bedrock deformation} model used in this thesis describes the flexure of
an elastic
lithosphere on top of an infinite half-space viscous mantle
\citep{Lingle.Clark.1985}. It can be described by a single pseudo-differential
equation of the bed elevation, $b$,
\begin{equation}
    \label{eqn:beddef}
    2\nu_m\,|\grad|\,\pdv{b}{t} + \rho_l g b + D\,\Delta^2 b = \sigma_{zzb} \,,
\end{equation}
where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
lithosphere's flexural rigidity (\cref{tab:bedrockparams}), and $\sigma_{zzb}$
corresponds to the ice load
\citep{Bueler.etal.2007}. Here, ${\Delta^2 = (\div\grad{})^2}$ designs the
biharmonic (Laplacian square) operator of
linear elastic theory, while $|\grad{}|$ is a pseudo-differential operator
defined through its Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. On
the left-hand side of \cref{eqn:beddef}, the first term accounts for mantle
relaxation, the second for point-wise isostasy, and the third for elastic
flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
the ice sheet growth and the isostatic bedrock response.

@@

In the simulations presented here, \cref{eqn:calovgreve} is numerically
approximated using week-long sub-intervals. For each sub-interval, PDDs, which
serve as a proxy for the available energy for melt, define the amount of snow
melt through a proportionality factor $F_s$. If all snow is melted, remaining
PDDs are then used to melt ice, using a proportionality factor $F_i$
(\cref{fig:model-pdd}). The degree-day factors for snow, $F_s$, and ice, $F_i$
(\cref{tab:surfaceparams}),
are derived from mass-balance measurements on contemporary glaciers in the
Coast and Rocky mountains of British Columbia \citep{Shea.etal.2009}. In
{\CCLI}, the temperature standard deviation $\sPDD$ is a constant model
parameter. In {\CCYC}, it is computed from daily temperature values from the
North American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, using
a method initially developed in {\PSDV}, and further improved upon in {\PSDP}.
Although the PDD model is a crude parametrisation of processes happening at the
surface of a glacier, its applicability has been proven \citep{Hock.2003},
while other, physically-based methods \citep[e.g.][]{Hock.2005}, require a
precise knowledge of the physical state of the atmosphere, typically very
uncertain for past periods such as the last glacial cycle.

@@

In \CCLI, we test the sensitivity of the ice sheet model to five different
input present-day climatologies, $\{T_{m0}, P_{m0}\}$ (\cref{tab:datasources}),
while adopting a range of constant temperature-offset corrections,
${\Delta}T_{TS}$, between -10 and 0\,K. In \CCYC, on the opposite, we test ice
sheet model sensitivity to time dependent temperature-offset corrections
derived from six different palaeo-temperature proxy records
(\cref{tab:datasources}).

>>>>

The North American Cordillera is an immense maze of mountain ranges and
intermontane basins which once hosted an ice mass of the size of Greenland: the
Cordilleran ice sheet. Although less extensive than the former Laurentide and
Fennoscandian ice sheets, the Cordilleran ice sheet perhaps remains the least
understood Pleistocene ice sheet in terms of its extent, volume, and dynamics.
Because of the irregular topography of the ice sheet's subglacial landscape,
ice flowed along tortuous pathways, sometimes following the major topographical
troughs, and sometimes flowing across them.

@@

The simulations presented in this thesis demonstrate the potential of combining
numerical ice sheet modelling to analysis of geological evidence in order to
improve our understanding of the former glaciations.

@@

Nevertheless, the landform record is sparse in time and, most often, spatially
incomplete. Palaeo-ice sheets did not leave a continuous imprint on the
lanscape, and much of this evidence has been overprinted by subsequent glacier
re-advances and other geomorphological processes \citep{Kleman.1994,
Kleman.etal.2006, Kleman.etal.2010}.
Therefore, geomorphological reconstructions of former glacier extent require a
substantial amount of interpretation. This is where numerical ice sheet models,
incorporating the processes of glacier sliding \citep[e.g.,][]{Weertman.1957}
and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can
help. Although these models alone contain too many uncertainties to reproduce
accurate ice sheet reconstructions during former glacial cycles
\citep[e.g.,][]{Hebeler.etal.2008}, they can be calibrated against geological
data where they exist \citep{Napieralski.etal.2007}, while allowing
physics-based, quantitative reconstructions of ice sheet sectors where they do
not
\citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.
In addition, numerical ice sheet models provide quantitative insights on the
four-dimensional distribution of stresses and deformations in palaeo-ice sheets
and allow for reconstructions of their past thickness and volume evolution, which
are difficult to infer from geological evidence\renote{
    Is this the place where to cite \citep{Denton.Hughes.1981}? \todo{I need to
    check the contents of this book (library).}
}.

@@

The \emph{dominant reliefs} occur along two major ribbons that run parallel to
each other along the general orientation of the North American Cordillera. The
western ribbon is formed by the Pacific Coast orogenic belt and includes, from
north to south, the Alaska Range (\cref{fig:photo-alaska-range}), the Wrangell
and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern ribbon consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench depression, and includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky mountains in the south. The
Brooks Range of northern Alaska has been covered by ice, but its glaciers
remained disjunct from the main body of the Cordilleran ice sheet
\citep{Kaufman.Manley.2004}.

@@

Evidence from glaciofluvial deposits dated using cosmogenic nuclides at
$2.64^{+0.20}/_{-0.18}$\,Ma indicates the first presence of an ice sheet in the
Cordilleran mountains during the late Pliocene \citep{Hidy.etal.2013}.
This indicates that the North American Cordillera has been glaciated many times
since the late Pliocene, with the first of these glaciations occurring even
before the mid-Pleistocene onset of 100 ka glacial cycles
(\cref{fig:paleo-timeseries}). However, only fragments of the last few
glaciations can be reconstructed from the geologic record, implying that much
of the older evidence has been erased by subsequent glacial re-advances and
perhaps even by other, non-glacial geomorphological processes.

@@

Apart from geological evidence of the final demise of the Cordilleran ice sheet, most field
evidence relates to its LGM extent. It indicates that ice covered most of the
Canadian Cordillera, and some of the adjacent regions in north-western
contiguous USA and Alaska (\cref{fig:map-cordillera}). The Cordilleran ice
sheet originated from the coalescence of several mountain-centred ice caps
\citep{Davis.Mathews.1944}. It is generally thought that this LGM configuration
was short-lived \citep{Clague.etal.1980, Clague.1985, Cosma.etal.2008}, a
hypothesis which can be studied through numerical modelling.

The \emph{southern margin} of the ice sheet has been studied extensively, and
is in several respects better understood than its northern counterpart
\citep{Booth.etal.2003}. It consists of coalescent piedmont lobes, each of
which formed at the outlet of the main valleys. The Puget Lobe
\citep{Thorson.1980, Porter.Swanson.1998} was the westernmost and largest one.
It covered the Puget Lowland, a region that comprises the cities of Vancouver
and Seattle. Further east, beyond the Cascade Range, the Okanagan and Purcell
lobes formed the southern Cordilleran ice sheet margin. The latter dammed
the Clark Fork River, inhibiting runoff and forming Glacial Lake Missoula
\citep{Pardee.1910}. As the ice sheet margin retreated, the lake repeatedly and
catastrophically drained to the Pacific Ocean \citep{Bretz.1923,Waitt.1980}.
These floods contributed to shape a distinctive landscape, known as the
Channeled Scabland.

The \emph{western margin} of the Cordilleran ice sheet covered part of the
Pacific continental shelf. An independent ice cap covering Vancouver Island
diverted flow from the main body of ice to drain into the Strait of Georgia
between Vancouver Island and the Olympic Peninsula, where it formed the Juan
de Fuca marine ice lobe, calving into the Pacific Ocean \mref. Further to the
North, the ice margin encroached the Queen Charlotte Islands. However, the
archipelago may not have been fully ice-covered, even during the maximum of
glaciation. Submarine troughs traversing several portions of the continental
shelf indicate the erosional action of additional ice streams, yet it is not
clear whether they have formed during the last glacial cycle or during earlier
glaciations \mref.

@@

The retreat of the ice sheet during the last \emph{deglaciation} was likely
rapid. Most of the evidence for the style and pace of deglaciation comes from
the Fraser Plateau. Studies there have indicated that some of the highest
uplands emerged from the ice sheet prior to the surrounding lowlands, possibly
resulting in thin, stagnant ice remnants covering parts of the plateau
\citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}. Less
is known about the deglaciation of the northern sector of the Cordilleran ice
sheet, but abundant glacial landforms from the Liard Lowland indicate a
different mode of deglaciation, with active retreat of the ice margin towards
the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in
\CCYC\renote{
    I have not had time to dig into new literature. The Ward paper on
    deglacial valley fills in the Pelly River is like Chinese to me, and I
    could not access the Dyke paper on the Liard Lowland, assuming this is the
    one cited by Martin in his paper.}.

@@

The thermodynamic core of the ice sheet model relies on four elementary field
equations: the equation of incompressibility, the balance of stresses, the constitutive law for ice, and the conservation of energy. Firstly, ice flow is assumed \emph{incompressible}. Hence, ice
density, $\rho$, is constant, and
conservation of mass therefore implies a conservation of volume, which can be expressed
in terms of the ice velocity vector, $\vv$ (\cref{fig:model-variables}), by
\begin{equation}
    \label{eqn:incompressibility}
    \div{\vv} = 0 \,.
\end{equation}

@@

@@

Thirdly, after defining the strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$\renote{
  $\grad$, $\div$, $\tgrad$, $\tdiv$, are the classic gradient and divergence operators, applied on vector and tensor fields. $\tr$ is the trace of a matrix. I don't think these need to be defined.},
and the deviatoric stress tensor, ${\DST = \CST + p\,\IDT}$,
where $p=-\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
\citep{Nye.1953} based on laboratory experiments \citep{Glen.1952} is used to
relate these two quantities. It reads:
\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST \,.
\end{equation}
The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$, and
water content, $\omega$, through an Arrhenius-type law
\citep[Eqs.~63--65]{Paterson.Budd.1982, Aschwanden.etal.2012},
\begin{equation}
    A =
    \begin{cases}
        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
            & \text{if}\ T < T_c \,, \\
        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
            & \text{if}\ T \ge T_c \,,
    \end{cases}
\end{equation}
where $T_{pa}$ is the pressure-adjusted temperature calculated using the
Clapeyron relation, ${T_{pa} = T - \beta p}$. $R$ is the ideal gas constant,
and $A_c$, $A_w$, $Q_c$ and $Q_w$, are constant parameter corresponding to
values measured below and above a critical temperature threshold
${T_c=-10}$\,\degC\ \citep[\cref{tab:iceflowparams};][]{Paterson.Budd.1982}.
The water fraction, $\omega$, is capped at a maximum value of 0.01, above which
no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
Greve.1997}.

For a convenient derivation of the shallow shelf approximation, developed in
the next section (\cref{sec:siassa}), the constitutive law can be rewritten
using an inverted formulation. Assuming
${{\doteps_e}^2 = \frac{1}{2} \tr(\SRT^2)}$ and ${B=A^{-1/n}}$,
\cref{eqn:glenslaw} becomes
\begin{equation}
    \label{eqn:invglenslaw}
    \DST = B\,\doteps_e^{(1/n)-1}\,\SRT \,.
\end{equation}

By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
by ${\DST = 2 \nu\SRT}$, thus yielding
\begin{equation}
    \label{eqn:viscosity}
    \nu = \frac{B}{2}\,\doteps_e^{(1/n)-1} \,.
\end{equation}

@@

From \cref{eqn:vsia} it can be seen that shallow ice velocities, $\vsia$,
can be directly computed from the geometry of the ice sheet surface, $s$,
after a single vertical integration over the ice thickness, in order to
account for variations of ice rheology, $A$, with depth.

@@

In PISM, the SIA and SSA velocities are eventually added
\citep[\cref{fig:model-siassa};][Eq.~15]{Winkelmann.etal.2011}:
\begin{equation}
    \label{eqn:siassa}
    \vv = \vsia + \vssa \,.
\end{equation}
Because $\vsia$ is small where $\vssa$ is large, and
reciprocally, we can expect the model to perform well in regions where
conditions for applicability of either the SIA or the SSA are met
(\cref{fig:model-siassa}). However,
\cref{eqn:siassa} is a heuristic, and its validity in the zone of
transition between shear flow (SIA) and longitudinal flow (SSA) has not been
mathematically proven.

@@

The till cohesion $c_0$ is assumed to be zero\renote{
  Yes, we should probably defend this choice, but I am not aware of any
  observations justifying this.}.
The effective pressure on the
till, $N$, is determined by the modelled amount of water at the bed. However,
two different parametrisations are used within this thesis. In {\CCLI}, using
PISM~0.5, the effective pressure is linearly related to water content by a
simple parametrisation \citep[Eq.~13]{Winkelmann.etal.2011},
\begin{equation}
    N = \rho gh (1 - \alpha \frac{W}{W_{max}}) \,,
\end{equation}
where $W$ is the amount of water at the bed, and varies from zero to
$W_{max}=2$\,m, a threshold above which further melt simply exits the system,
thus assuming instantaneous drainage. Although mass-conserving water drainage
is now implemented in PISM \citep{Bueler.Pelt.2014}, it is not included in the
simulations presented here for the sake of simplicity.

@@

The modelled ice sheet interacts with the underlying bedrock in two ways.
Firstly, \emph{bedrock temperatures} are computed to a given depth by
appending the
ice enthalpy model (\ref{eqn:enthalpy}) with an underlying bedrock thermal
model. The only process accounted for is vertical heat conduction, hence this model
consists of a simple diffusion equation,
\begin{equation}
    \pdv{T}{t} = \frac{k_b}{\rho_b c_b} \Delta T \,,
\end{equation}
where $k_b$, $\rho_b$ and $c_b$ are the thermal conductivity, density and
specific heat capacity of the bedrock (\cref{tab:bedrockparams}).
Bedrock temperature is conditioned at depth by a fixed upwards geothermal heat
flux $q_G$  (\cref{tab:bedrockparams}). The bedrock thermal model
is necessary for modelling the Cordilleran ice sheet on multi-millennial
time-scales, because temperature fluctuations caused by climate change and
varying degrees of isolation from atmospheric temperatures by ice sheet
coverage, penetrate several kilometres
into the rock, and eventually affect basal ice temperatures.

Secondly, bedrock elevation evolves in response to the ice weight. The
\emph{bedrock deformation} model used in this thesis describes the flexure of
an elastic
lithosphere on top of an infinite half-space viscous mantle
\citep{Lingle.Clark.1985}. It can be described by a single pseudo-differential
equation of the bed elevation, $b$,
\begin{equation}
    \label{eqn:beddef}
    2\nu_m\,|\grad|\,\pdv{b}{t} + \rho_l g b + D\,\Delta^2 b = \sigma_{zzb} \,,
\end{equation}
where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
lithosphere's flexural rigidity (\cref{tab:bedrockparams}), and $\sigma_{zzb}$
corresponds to the ice load
\citep{Bueler.etal.2007}. Here, ${\Delta^2 = (\div\grad{})^2}$ designs the
biharmonic (Laplacian square) operator of
linear elastic theory, while $|\grad{}|$ is a pseudo-differential operator
defined through its Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. On
the left-hand side of \cref{eqn:beddef}, the first term accounts for mantle
relaxation, the second for point-wise isostasy, and the third for elastic
flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
the ice sheet growth and the isostatic bedrock response.

@@

In the simulations presented here, \cref{eqn:calovgreve} is numerically
approximated using week-long sub-intervals. For each sub-interval, PDDs, which
serve as a proxy for the available energy for melt, define the amount of snow
melt through a proportionality factor $F_s$. If all snow is melted, remaining
PDDs are then used to melt ice, using a proportionality factor $F_i$
(\cref{fig:model-pdd}). The degree-day factors for snow, $F_s$, and ice, $F_i$
(\cref{tab:surfaceparams}),
are derived from mass-balance measurements on contemporary glaciers in the
Coast and Rocky mountains of British Columbia \citep{Shea.etal.2009}. In
{\CCLI}, the temperature standard deviation $\sPDD$ is a constant model
parameter. In {\CCYC}, it is computed from daily temperature values from the
North American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, using
a method initially developed in {\PSDV}, and further improved upon in {\PSDP}.
Although the PDD model is a crude parametrisation of processes happening at the
surface of a glacier, its applicability has been demonstrated
\citep[e.g.][]{Hock.2003},
while other, physically-based methods \citep[e.g.][]{Hock.2005}, require a
precise knowledge of the physical state of the atmosphere, which typically is highly
uncertain for past periods such as the last glacial cycle.

@@

In \CCLI, we test the sensitivity of the ice sheet model to five different
input present-day climatologies, $\{T_{m0}, P_{m0}\}$ (\cref{tab:datasources}),
while adopting a range of constant temperature-offset corrections,
${\Delta}T_{TS}$, between -10 and 0\,K. In \CCYC, on the contrary, we test ice
sheet model sensitivity to time dependent temperature-offset corrections
derived from six different palaeo-temperature proxy records
(\cref{tab:datasources}).
